The human visual system preserves the hierarchy of 2-dimensional pattern regularity

### 2021-02-03 Writeup/manuscript.tex
Figure 1 text wrap
less critical discussion of correlation

% Please add here a significance statement to explain the relevance of your work
\significancestatement{Wallpaper groups were discovered in the mid-19th century, and the 17 groups constitute the complete set of possible ways of regularly tiling the 2D-plane. In recent years wallpaper groups have found use in the vision science community, as an ideal stimulus set for studying the perception of symmetries in textures. Here we present brain imaging and psychophysical data using the complete set of wallpaper groups and show that the hierarchical organization among wallpaper groups is reflected both in representations in visual cortex and performance on a symmetry detection task. This shows that the visual system is highly sensitive to regularities in textures, and suggest that symmetries may play an important role in texture perception.}

@@

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\dropcap{S}ymmetries are present at many scales in images of natural scenes, due to a complex interplay of physical forces that govern pattern formation in nature. The importance of symmetry for visual perception has been known at least since the gestalt movement of the early 20th century. Since then, symmetry has been shown to contribute to the perception of shapes \cite{RN1311,RN1682}, scenes \cite{RN1824} and surface properties \cite{RN1166}, as well as the social process of mate selection \cite{RN1337}. Most of this work has focused on mirror symmetry or \textit{reflection}, with much less attention being paid to the other fundamental symmetries: \textit{rotation}, \textit{translation} and \textit{glide reflection}. In the two spatial dimensions relevant for images, these four fundamental symmetries can be combined in 17 distinct ways, \textit{the wallpaper groups} \cite{RN1562,RN1563,RN1425}. Previous work on a subset of the wallpaper groups used functional MRI to demonstrate that rotation symmetries in wallpapers elicit parametric responses in several areas in occipital cortex, beginning with visual area V3 \cite{RN1725}. This effect was also robust with electroencephalography (EEG), whether measured using Steady-State Visual Evoked Potentials (SSVEPs)\cite{RN1725} or event-related paradigms \cite{RN1959}. Here we extend this work by collecting SSVEPs and psychophysical data from human participants viewing the full set of wallpaper groups. We measure responses in visual cortex to 16 out of the 17 wallpaper groups, with the 17th serving as a control stimulus. Our goal is to provide a more complete picture of how wallpaper groups are represented in the human visual system.

@@

\begin{figure}[ptb]
\centering
\includegraphics[width=1\linewidth]{../figures/figure1.png}
\caption{Examples of each of the 16 wallpaper groups are shown in the left- and right-most column of the figures, next to the corresponding SSVEP (center-left) and psychological (center-right) data from each group. The SSVEP data are odd-harmonic-filtered cycle-average waveforms. In each cycle, a \textit{P1} exemplar was shown for the first 600 ms, followed by the original exemplar for the last 600 ms. Errorbars are standard error of the mean. Psychophysical data are presented as boxplots reflecting the distribution of display duration thresholds. The 16 groups are ordered by the strength of the SSVEP response, to highlight the range of response amplitudes.}
\label{fig:example_stimuli}
\end{figure}

@@

The psychophysical experiment took a distinct but related approach. In each trial an exemplar image was shown with its matched control, one image after the other, and the order varied pseudo-randomly such that in half the trials the original exemplar was shown first, and in the other half the control image was shown first. After each trial, participants were told to indicate whether the first or second image contained more structure, and the duration of both images was controlled by a staircase procedure so that a threshold duration for symmetry detection could be computed for each wallpaper group.

Examples of the wallpaper groups and a summary of our brain imaging and psychophysical measurements is in Figure 1. For our primary SSVEP analysis, we only considered EEG data from a pre-determined region-of-interest (ROI) consisting of six electrodes over occipital cortex (see Supplementary Figure 1.1). SSVEP data from this ROI was filtered so that only the odd harmonics that capture the symmetry response contribute to the waveforms. While waveform amplitude is quite variable among the 16 groups, all groups have a sustained negative-going response that begins at about the same time for all groups, ~180 ms after the transition from the \textit{P1} control exemplar to the original exemplar. To reduce the amplitude of the symmetry-specific response to a single number that could be used in further analyses and compared to the psychophysical data, we computed the root-mean-square (RMS) over the odd-harmonic-filtered waveforms. The data in Figure 1 are shown in descending order according to RMS. The psychophysical results, shown in box plots in Figure 1, were also quite variable between groups, and there seems to be a general pattern where wallpaper groups near the top of the figure, that have lower SSVEP amplitudes, also have longer psychophysical threshold durations.

@@

Finally, we conducted a correlation analysis comparing SSVEP and psychophysical data, and found a modest ($R^2 = 0.44$) but above-zero correlation. The correlation between the two datasets is interesting because it suggests that our psychophysical and SSVEP measurements are at least to some extent tapping into the same underlying mechanisms.

\begin{figure}[tbh]
\centering
\includegraphics[width=1\linewidth]{../analysis/plots/subgroup_comp_eeg_rms.png}
\caption{Posterior distributions for the difference in mean SSVEP RMS amplitude. Colour coding of the text indicates the index of the subgroup, while the colour of the filled distribution relates to the conditional probability that the difference in means is greater than zero. We can see that 56/63 subgroup relationships have $p(\Delta|data)>0.95$.}
\label{fig:eeg_comp}
\end{figure}

\begin{figure}[tbh]
\centering
\includegraphics[width=1\linewidth]{../analysis/plots/subgroup_comp_psychophysical.png}
\caption{Posterior distributions for the difference in mean symmetry detection threshold durations. Colour coding of the text indicates the index of the subgroup, while the colour of the filled distribution relates to the conditional probability that the difference in means is greater than zero. We can see that 49/63 subgroup relationships have $p(\Delta|data)>0.95$.}
\label{fig:durations_comp}
\end{figure}

@@

Here we show that beyond merely responding to the elementary symmetry operations of reflection \cite{RN1170} and rotation \cite{RN1725}, the visual system explicitly represents hierarchical structure of the 17 wallpaper groups, and thus every composition of the four fundamental symmetries (rotation, reflection, translation, glide reflection) which comprise the set of regular textures. Both SSVEP amplitudes and symmetry detection thresholds preserve the hierarchy of complexity among the wallpaper groups that is captured by the subgroup relationships \cite{RN1711}. For the SSVEP, this remarkable consistency was specific to the odd harmonics of the stimulus frequency that are known to capture the symmetry-specific response \cite{RN1725} and to electrodes in a region-of-interest (ROI) over occipital cortex. When the same analysis was done using the odd harmonics from electrodes over parietal cortex (Supplementary Figure 3.3) or even harmonics from electrodes over occipital cortex (Supplementary Figure 3.4), the data was substantially less consistent with the subgroup relationships (yellow and green lines, Figure 4).

The current data provide a complete description of the visual system’s response to symmetries in the 2-D plane. Our design precludes us from independently measure the response to \textit{P1}, but because each of the 16 other groups produce non-zero odd harmonic amplitudes (see Figure 1), we can conclude that the relationships between P1 and all other groups, where \textit{P1} is the subgroup, are also preserved by the visual system. The subgroup relationships are in many cases not obvious perceptually, and most participants had no knowledge of group theory. Thus, the visual system’s ability to preserve the subgroup hierarchy does not depend on explicit knowledge of the relationships. Previous behavioral experiments have shown that although naïve observers can distinguish many of the wallpaper groups \cite{RN1253}, they tend to sort them into fewer groups than there actually are (4-12 groups) and it is common for exemplars from different wallpaper groups to be sorted in the same group \cite{RN172}. The more controlled two-interval forced choice approach used in the current behavioral experiment allows us to show that more granular representations of wallpaper groups are measurable in behavior. The relatively modest correlation we observe between behavior and brain imaging data may suggest that the relationship between the two measurements is more complex than what can be captured in our summary statistics. Future work in which the same individuals participate in both behavioral and brain imaging experiments will help tease apart these complexities and enhance our understanding of the brain mechanisms that drive both measurements.

@@

The 17 wallpaper groups are completely regular, and relatively rare in the visual environment, especially when considering distortions due to perspective and occlusion. Near-regular textures, however, abound in the visual world, and can be approximated as deformed versions of the wallpaper groups \cite{RN1519}. The correspondence between visual cortex responses and group theory demonstrated here may indicate that the visual system represents visual textures using a similar scheme, with the wallpaper groups serving as anchor points in representational space. This framework resembles norm-based encoding strategies that have been proposed for other stimulus classes, most notably faces \cite{RN435}, and leads to the prediction that adaptation to wallpaper patterns should distort perception of near-regular textures, similar to the aftereffects found for faces \cite{RN1768}. Field biologists have demonstrated that animals respond more strongly to exaggerated versions of a learned stimulus, referred to as “supernormal” stimuli \cite{RN1775}. In the norm-based encoding framework, wallpaper groups can be considered \textit{supertextures}, exaggerated examples of the near-regular textures that surround us. Artists may consciously or unconsciously create supernormal stimuli, to capture the essence of the subject and evoke strong responses in the audience \cite{RN1764}. Wallpaper groups are visually compelling and have been widely used in human artistic expression going back to the Neolithic age \cite{RN1949}. If wallpapers are in fact supertextures, this prevalence may be a direct result of the strategy the human visual system has adopted for texture encoding.

>>>>

% Please add here a significance statement to explain the relevance of your work
\significancestatement{Wallpaper groups were discovered in the mid-19th century, and the 17 groups constitute the complete set of possible ways of regularly tiling the 2-D plane. In recent years wallpaper groups have found use in the vision science community, as an ideal stimulus set for studying the perception of symmetries in textures. Here we present brain imaging and psychophysical data using the complete set of wallpaper groups and show that the hierarchical organization among wallpaper groups is reflected both in representations in visual cortex and performance on a symmetry detection task. This shows that the visual system is highly sensitive to regularities in textures, and suggest that symmetries may play an important role in texture perception.}

@@

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\dropcap{S}ymmetries are present at many scales in natural and man-made environments, due to a complex interplay of physical forces that govern pattern formation in nature. Symmetrical patterns have been created and appreciated by human cultures throughout history and since the gestalt movement of the early 20th century, symmetry has been recognized as important for visual perception. Symmetry contributes to the perception of shapes \cite{RN1311,RN1682}, scenes \cite{RN1824} and surface properties \cite{RN1166}, as well as the social process of mate selection \cite{RN1337}. Most of this work has focused on mirror symmetry or \textit{reflection}, with much less attention being paid to the other fundamental symmetries: \textit{rotation}, \textit{translation} and \textit{glide reflection}. In the two spatial dimensions relevant for images, these four symmetries can be combined in 17 distinct ways, \textit{the wallpaper groups} \cite{RN1562,RN1563,RN1425}. Previous work on a subset of four of the wallpaper groups used functional MRI to demonstrate that rotation symmetries in wallpapers elicit parametric responses in several areas in occipital cortex, beginning with visual area V3 \cite{RN1725}. This effect was also robust with electroencephalography (EEG), whether measured using Steady-State Visual Evoked Potentials (SSVEPs)\cite{RN1725} or event-related paradigms \cite{RN1959}. Here we extend this work by collecting SSVEPs and psychophysical data from human participants viewing the full set of wallpaper groups. We measure responses in visual cortex to 16 out of the 17 wallpaper groups, with the 17th serving as a control stimulus. Our goal is to provide a more complete picture of how wallpaper groups are represented in the human visual system.

@@

\begin{figure}[h!tb]
\includegraphics[width=1\linewidth]{../figures/figure1.png}
\caption{Examples of each of the 16 wallpaper groups are shown in the left- and right-most column of the figures, next to the corresponding SSVEP (center-left) and psychological (center-right) data from each group. The SSVEP data are odd-harmonic-filtered cycle-average waveforms. In each cycle, a \textit{P1} exemplar was shown for the first 600 ms, followed by the original exemplar for the last 600 ms. Errorbars are standard error of the mean. Psychophysical data are presented as boxplots reflecting the distribution of display duration thresholds. The 16 groups are ordered by the strength of the SSVEP response, to highlight the range of response amplitudes.}
\label{fig:example_stimuli}
\end{figure}

@@

The psychophysical experiment took a distinct but related approach. In each trial an exemplar image was shown with its matched control, one image after the other, and the order varied pseudo-randomly such that in half the trials the original exemplar was shown first, and in the other half the control image was shown first. After each trial, participants were instructed to indicate whether the first or second image contained more structure. The duration of both images was controlled by a staircase procedure so that a threshold duration for symmetry detection could be computed for each wallpaper group.

Examples of the wallpaper groups and a summary of our brain imaging and psychophysical measurements are shown in Figure 1. For our primary SSVEP analysis, we only considered EEG data from a pre-determined region-of-interest (ROI) consisting of six electrodes over occipital cortex (see Supplementary Figure 1.1). SSVEP data from this ROI was filtered so that only the odd harmonics that capture the symmetry response contribute to the waveforms. While waveform amplitude is quite variable among the 16 groups, all groups have a sustained negative-going response that begins at about the same time for all groups, ~180 ms after the transition from the \textit{P1} control exemplar to the original exemplar. To reduce the amplitude of the symmetry-specific response to a single number that could be used in further analyses and compared to the psychophysical data, we computed the root-mean-square (RMS) over the odd-harmonic-filtered waveforms. The data in Figure 1 are shown in descending order according to RMS. The psychophysical results, shown in box plots in Figure 1, were also quite variable between groups, and there seems to be a general pattern where wallpaper groups near the top of the figure, that have lower SSVEP amplitudes, also have longer psychophysical threshold durations.

@@

Finally, we conducted a correlation analysis comparing SSVEP and psychophysical data, and found a reliable correlation ($R^2 = 0.44$, Bayesian confidence interval [0.28, 0.55]). The correlation reflects an inverse relationship: For subgroup relationships where the supergroup produces a much \textit{larger} SSVEP amplitude than the subgroup, the supergroup also tends to produce a much \textit{smaller} symmetry detection threshold. This is consistent with our hypotheses about how the two measurements relate to symmetry representations in the brain, and suggests that our brain imaging and psychophysical measurements are at least to some extent tapping into the same underlying mechanisms.

\begin{figure}[tbh]
\centering
\includegraphics[width=1\linewidth]{../analysis/plots/subgroup_comp_eeg_rms.png}
\caption{Posterior distributions for the difference in mean SSVEP RMS amplitude. Colour coding of the text indicates the index of the subgroup, while the colour of the filled distribution relates to the conditional probability that the difference in means is greater than zero. We can see that 55/63 subgroup relationships have $p(\Delta|data)>0.99$.}
\label{fig:eeg_comp}
\end{figure}

\begin{figure}[tbh]
\centering
\includegraphics[width=1\linewidth]{../analysis/plots/subgroup_comp_psychophysical.png}
\caption{Posterior distributions for the difference in mean symmetry detection threshold durations. Colour coding of the text indicates the index of the subgroup, while the colour of the filled distribution relates to the conditional probability that the difference in means is greater than zero. We can see that 43/63 subgroup relationships have $p(\Delta|data)>0.99$.}
\label{fig:durations_comp}
\end{figure}

@@

Here we show that beyond merely responding to the elementary symmetry operations of reflection \cite{RN1170} and rotation \cite{RN1725}, the visual system represents the hierarchical structure of the 17 wallpaper groups, and thus every composition of the four fundamental symmetries (rotation, reflection, translation, glide reflection) which comprise the set of regular textures. Both SSVEP amplitudes and symmetry detection thresholds preserve the hierarchy of complexity among the wallpaper groups that is captured by the subgroup relationships \cite{RN1711}. For the SSVEP, this remarkable consistency was specific to the odd harmonics of the stimulus frequency that are known to capture the symmetry-specific response \cite{RN1725} and to electrodes in a region-of-interest (ROI) over occipital cortex. When the same analysis was done using the odd harmonics from electrodes over parietal cortex (Supplementary Figure 3.3) or even harmonics from electrodes over occipital cortex (Supplementary Figure 3.4), the data was substantially less consistent with the subgroup relationships (yellow and green lines, Figure 4).

The current data provide a description of the visual system’s response to the complete set of symmetries in the two-dimensional plane. Our design precludes us from independently measure the response to \textit{P1}, but because each of the 16 other groups produce non-zero odd harmonic amplitudes (see Figure 1), we can conclude that the relationships between P1 and all other groups, where \textit{P1} is the subgroup, are also preserved by the visual system. The subgroup relationships are in many cases not obvious perceptually, and most participants had no knowledge of group theory. Thus, the visual system’s ability to preserve the subgroup hierarchy does not depend on explicit knowledge of the relationships. Previous behavioral experiments have shown that although naïve observers can distinguish many of the wallpaper groups \cite{RN1253}, they tend to sort them into fewer groups than there actually are (4-12 groups) and it is common for exemplars from different wallpaper groups to be sorted in the same group \cite{RN172}. The more controlled two-interval forced choice approach used in the current behavioral experiment allows us to show that more granular representations of wallpaper groups are measurable in behavior.

We observe a reliable correlation between our brain imaging and psychophysical data. This suggests that the two measurements reflect the same underlying symmetry representations in visual cortex. While it should be noted that the correlation is relatively modest ($R^2 = 0.44$), we note that this may be partly due to the fact that the same individuals did not participate in the two experiments. Future work in which behavioral and brain imaging data are collected from the same participants, will help further establish the connection between the two measurements, and tease apart any additional complexity that may not have been captured by the summary statistics we applied here.

@@

The 17 wallpaper groups are completely regular, and relatively rare in the visual environment, especially when considering distortions due to perspective and occlusion. Near-regular textures, however, abound in the visual world, and can be approximated as deformed versions of the wallpaper groups \cite{RN1519}. The correspondence between visual cortex responses and group theory demonstrated here may indicate that the visual system represents visual textures using a similar scheme, with the wallpaper groups serving as anchor points in representational space. This framework resembles norm-based encoding strategies that have been proposed for other stimulus classes, most notably faces \cite{RN435}, and leads to the prediction that adaptation to wallpaper patterns should distort perception of near-regular textures, similar to the aftereffects found for faces \cite{RN1768}. Field biologists have demonstrated that animals respond more strongly to exaggerated versions of a learned stimulus, referred to as “supernormal” stimuli \cite{RN1775}. In the norm-based encoding framework, wallpaper groups can be considered \textit{supertextures}, exaggerated examples of the near-regular textures that surround us. Artists may consciously or unconsciously create supernormal stimuli, to capture the essence of the subject and evoke strong responses in the audience \cite{RN1764}. Wallpaper groups are visually compelling and symmetries have been widely used in human artistic expression going back to the Neolithic age \cite{RN1949}. If wallpapers are in fact supertextures, this prevalence may be a direct result of the strategy the human visual system has adopted for texture encoding.
